Robert C. Dunnell’s _Systematics in Prehistory_ at 50

### 2022-04-21 analysis/paper/paper.Rmd
merge in some edits on a docx that I missed previously

---
title: "Robert C. Dunnell’s _Systematics in Prehistory_ at 50"
author:
  - Felix Riede:
      email: f.riede@cas.au.dk
      institute: [UofO]
      correspondence: true
  - Astolfo Araujo:
      institute: [UofSP]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofWA]
      correspondence: false
institute:
  - UofO: Department of Archaeology and Heritage Studies, Aarhus University, Moesgård Allé 20, 8270 Højbjerg, Denmark
  - UofSP: Museum of Archaeology and Ethnology, University of São Paulo, Brazil
  - UofWA: Department of Anthropology, University of Washington, Seattle, WA, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
       - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/apa.csl" # Insert path for the bib-sty
abstract: |
  2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell’s (1971) diminutive yet dense _Systematics in Prehistory_. At the height of the debate between Culture History and New Archaeology, Dunnell’s work sought to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics. Dunnell himself was notorious and controversial, however, but the importance of his work remains underappreciated. Like other precocious works of that tumultuous time Systematics in Prehistory today remains absent from most course reading lists and gathers dust on library shelves. In this contribution we argue for a greater appreciation of its as yet unfulfilled conceptual and analytical promise. In particular, we briefly chart its somewhat delayed impact via evolutionary archaeology, including how it has also influenced non-Anglophone traditions, especially in South America. The obstinate persistence of classification issues in paleoanthropology and Paleolithic archaeology, we argue, warrants a second look at Dunnell's _Systematics_.
---

@@

Evolutionary Archaeology is concerned with the study of cultural evolution through material culture proxies and commonly over extended timescales not readily accessible by other means. While evolutionary thinking *per se* is not new in the discipline, an operational evolutionary archaeology emerged only recently, and on the back of several major paradigm transitions [@prentiss2019handbook]. The notion of major paradigm shifts in the history of archaeology is, however, characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s do stand out as revolutionary. In the US, the Binfordian juggernaut increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's [@clarke1968a] formidable *Analytical Archaeology* was shaking up the establishment with novel concepts and methods as well as incisive rhetoric. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers working within ecological and evolutionary approaches to past culture change are rediscovering the merits of Clarke's conceptual approach in particular, and Binford's extensive data synthesis [@nicholas2012; @lycett2018a]. The year 2021 marked the fiftieth anniversary of the publication of Robert C. Dunnell's [@dunnell1971] diminutive yet dense *Systematics in Prehistory*, a volume that is concerned entirely with classification in archaeology. Appearing at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its goal of addressing a more fundamental issue that was and still is relevant to all branches of archaeology: systematics.

Dunnell himself was notorious and controversial, and his *Systematics* was received with mixed reactions by his contemporaries who commented - mostly negatively - both on his style of writing and the book's content [@spaulding1974; @tuggle1974; @bayard1973; @shenkel1973]. Like other precocious works of that tumultuous time *Systematics in Prehistory* today remains absent from most course reading lists and gathers dust on library shelves. As Lyman has recently shown, systematics takes up little space in contemporary archaeological research or teaching [@lyman2021]. In striking contrast, biological systematics is a well developed field with faculty positions, courses, and journals dedicated to this fundamental scientific concern. If at least some of the success of Clarke and Binford can be attributed to them boldly tackling exciting and large-scale topics such as migration and adaptation adaptation through the introduction of avant-garde terminology, then the corresponding obscurity of Dunnell's *Systematics* can perhaps be attributed to him focusing on an issue that simply seemed too quotidian. Furthermore, archaeology has moved ahead in such a way that leaves some of the key claims *Systematics* exposed as a distinct outliers in modern American archaeology. For example, his insistence that the exclusive focus of archaeology be the physical traces of pasts human activity, and that archaeology cannot be both a science and a sub-discipline of anthropology, are generally minority positions now, with archaeology continuing to be a sub-field of anthropology in most US universities at least. In countries where archaeology is situated in different institutional contexts, this particular issue is not a major concern. Be it as it may, Dunnell was correct in his axiomatic insistence that rigorous classification comes before any other analysis or interpretation. Without consistent and explicit classification, any scientific discipline will inevitably fail to produce cumulative insights; certainly, evolutionary analysis would hardly be possible without robust classification.

@@

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary deep history: the Palaeolithic. Rooted in French antiquarianism, the development of Paleolithic systematics, for instance, has been likened to so many 'accidents of history' [@clark2009], but there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units [@reynolds2019]. The use of older typological classifications remains prevalent, despite clear and repeated critiques [e.g. @bisson2000]. More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches. However, theoretical explications of the generative mechanisms, and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell, remain exceptions rather than the rule [@tostevin2013]. Today, a great deal of attention is again being paid to systematics in prehistory, and as Barton and Clark [@barton2021] have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions. Several researchers are tackling classificatory issues with novel and mostly quantitative means [@ivanovait2020; @grove2021; @leplongeon2020]. At the same time, however, the topic remains poorly heeded in archaeological pedagogy [@lyman2021]. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching [@Riede_Walsh_Nowell_Langley_Johannsen_2021], we recommend that, after half a century, Dunnell's *Systematics in Prehistory* - and with it rigorous and replicable ways of classifying material culture - are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [-@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

```{r sip-citations-of-citing-work, fig.cap="Citations to 'Systematics' over time. Inset shows distributions of citations to works citing 'Systematics'. Data collected from Google Scholar on July 2021", fig.width = 5, fig.height = 5}
knitr::include_graphics(here::here("analysis/figures/citations-of-citing-works-over-time.png"))
```

```{r sip-keywords-in-titles, fig.cap="Keywords in titles of works citing 'Systematics'. Inset shows languages of works citing 'Systematics'", fig.width = 5, fig.height = 6}
knitr::include_graphics(here::here("analysis/figures/languages-and-keywords-in-titles-of-citing-works.png"))
```

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [@Cardillo_Charlin_2018], Mercedes Okumura [@Okumura_Araujo_2014] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. While the work by Cardillo and colleagues [-@Cardillo_Charlin_2018; -@cardillo2015evolution] focuses on phylogenetic signals and material culture diversity across time and space, Okumura and colleagues [-@Okumura_Araujo_2014] are more concerned with the persistence of attributes over long stretches of time. What these works have in common is the use of an explicitly cultural evolutionary rationale. In these studies, the use of paradigmatic classifications is widespread and an overt understanding of the difference between classes (ideational/theoretical) and groups (phenomenological/empirical) evident, underlining the influence of Dunnell’s thinking. Worth noting is that these approaches were embraced chiefly in only two South American countries – Argentina and Brazil – and largely independently of each other. Collectively, these works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields - first and foremost palaeontology - dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

```{r sip-topic-model, fig.cap="Left: Top twenty topics in titles of works citing 'Systematics'. The gamma value indicates their overall abundance, and the topics are labelled with the most heavily weighted words in each topic. Right upper: Clusters of citing works according to topic similarity, with clusters labelled by the most prominent topic. Clusters were computed by Principal Components Analysis of the topic proportions in each citing work, then a t-SNE to reduce dimensionality, and density based clustering and to identify clusters. Each data point is one document. Right lower: Plot of topics showing coherence and exclusivity metrics for each topic. Each data point is one topic", fig.width = 5, fig.height = 6}
knitr::include_graphics(here::here("analysis/figures/topic-model-on-titles.jpg"))
```

@@

The influence of *Systematics* has extended well beyond Dunnell's own primary study area of the North American Southeast. For example, the Portuguese translation published in 2007 has received 75 citations, and the Spanish translation 30. At face value these numbers seem low, but when we take into consideration that Binford's Portuguese translation of *In Pursuit of the Past* published in 1991 has 31 citations, we can have a better perspective of the influence Dunnell's work. On the other hand, in our data we see only a single work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that the book is quite difficult to read, as noted by reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [@spaulding1974 p. 515-516] described his reaction to the book), and acknowledged by Dunnell in his foreword to the 2002 edition. The low readability of the book has likely limited the accessibility of its contents for readers whose first language is not English, and made it inherently less attractive as a
pedagogical resource.

A second limitation of *Systematics* is that it failed to motivate the formation of a distinct research area and community of archaeological systematics. While biology has a Society of Systematic Biologists, and attendant journals such as *Systematic Biology*, *Systematics and Evolution*, and *Systematics and Biodiversity*, archaeology currently has no equivalent focal points for discussions of specifically culture-historical systematics that transcend specific geographical and chronological concerns. *Systematics* was not a work that served as an engine for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead, these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups. The result of this fragmentation is a proliferation of bespoke classifications that lack applicability across cases. It also inhibits reproducibility, synthesis, and the transfer of ideas and innovations across research communities, which may limit the sustainability of archaeology as a discipline.

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by the New Archaeology on the one hand, while also meeting resistance from deeply entrenched traditional approaches to classification. In addition, the volume also appeared just before personal computers were beginning to have a major impact on archaeological research practice [@clarkQuantificationAmericanArchaeology1982; @aldenderfer1987assessing]. The rigor advocated by Dunnell would have lent itself smoothly to the sorts of computational approaches now increasingly adopted in the discipline. Yet, the approaches of the New Archaeologists dominated the literature in the decade after the publication of *Systematics*, thoroughly eclipsing it as a discipline-defining text. Nevertheless, what makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by economies of personal prestige (Gabriel de Mortillet, 1821--1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets [cf. @Perreault_2019], the message of *Systematics* will only become more relevant.

@@

We thank Ruth Mace for encouraging us to submit this paper to Evolutionary Human Sciences. We also thank Mike O’Brien and one anonymous reviewer for the insightful comments on Dunnell and on our paper.

@@

We declare no conflict of interests.

@@

The data that support the findings of this study are openly available in the Open Science Framework at  <http://doi.org/10.17605/OSF.IO/JBPFW>. Readers may freely access these resources to reproduce all of the figures in the paper.

>>>>

---
title: "Robert C. Dunnell’s _Systematics in Prehistory_ at 50"
author:
  - Felix Riede:
      email: f.riede@cas.au.dk
      institute: [UofO]
      correspondence: true
  - Astolfo Araujo:
      institute: [UofSP]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofWA]
      correspondence: false
institute:
  - UofO: Department of Archaeology and Heritage Studies, Aarhus University, Moesgård Allé 20, 8270 Højbjerg, Denmark
  - UofSP: Museum of Archaeology and Ethnology, University of São Paulo, Brazil
  - UofWA: Department of Anthropology, University of Washington, Seattle, WA, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
       - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/apa.csl" # Insert path for the bib-sty
abstract: |
  2021 marked the fiftieth anniversary of the publication of Robert C. Dunnell’s (1971) diminutive yet dense _Systematics in Prehistory_. At the height of the debate between Culture History and New Archaeology, Dunnell’s work sought to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Palaeolithic: systematics. Dunnell himself was notorious and controversial, however, but the importance of his work remains underappreciated. Like other precocious works of that tumultuous time Systematics in Prehistory today remains absent from most course reading lists and gathers dust on library shelves. In this contribution we argue for a greater appreciation of its as yet unfulfilled conceptual and analytical promise. In particular, we briefly chart its somewhat delayed impact via evolutionary archaeology, including how it has also influenced non-Anglophone traditions, especially in South America. The obstinate persistence of classification issues in palaeoanthropology and palaeoarchaeology, we argue, warrants a second look at Dunnell's _Systematics_.
---

@@

Evolutionary Archaeology is concerned with the study of cultural evolution through material culture proxies and commonly over extended timescales not readily accessible by other means. While evolutionary thinking *per se* is not new in the discipline, an operational evolutionary archaeology emerged only recently, and on the back of several major paradigm transitions [@prentiss2019handbook]. The notion of major paradigm shifts in the history of archaeology is, however, characterized by a great deal of hyperbole and rhetorical manoeuvring That said, the late 1960s and 1970s do stand out as revolutionary. In the US, the Binfordian juggernaut increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's [-@clarke1968a] formidable *Analytical Archaeology* was shaking up the establishment with novel concepts and methods as well as incisive rhetoric. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers working within ecological and evolutionary approaches to past culture change are rediscovering the merits of Clarke's conceptual approach in particular, and Binford's extensive data synthesis [@nicholas2012; @lycett2018a]. The year 2021 marked the fiftieth anniversary of the publication of Robert C. Dunnell's [-@dunnell1971] diminutive yet dense *Systematics in Prehistory*, a volume that is concerned entirely with classification in archaeology. Appearing at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its goal of addressing a more fundamental issue that was and still is relevant to all branches of archaeology: systematics.

Dunnell himself was notorious and controversial, and his *Systematics* was received with mixed reactions by his contemporaries who commented—mostly negatively—both on his style of writing and the book's content [@spaulding1974; @tuggle1974; @bayard1973; @shenkel1973]. Like other precocious works of that tumultuous time *Systematics in Prehistory* today remains absent from most course reading lists and gathers dust on library shelves. As Lyman has recently shown, systematics takes up little space in contemporary archaeological research or teaching [@lyman2021]. In striking contrast, biological systematics is a well-developed field with faculty positions, courses, and journals dedicated to this fundamental scientific concern. If at least some of the success of Clarke and Binford can be attributed to them boldly tackling exciting and large-scale topics such as migration and adaptation through the introduction of avant-garde terminology, then the corresponding obscurity of Dunnell's *Systematics* can perhaps be attributed to him focusing on an issue that simply seemed too quotidian. Furthermore, archaeology has moved ahead in such a way that leaves some of the key claims *Systematics* exposed as distinct outliers in modern American archaeology. For example, his insistence that the exclusive focus of archaeology be the physical traces of pasts human activity, and that archaeology cannot be both a science and a sub-discipline of anthropology, are generally minority positions now, with archaeology continuing to be a sub-field of anthropology in most US universities at least. In countries where archaeology is situated in different institutional contexts, this particular issue is not a major concern. Be it as it may, Dunnell was correct in his axiomatic insistence that rigorous classification comes before any other analysis or interpretation. Without consistent and explicit classification, any scientific discipline will inevitably fail to produce cumulative insights; certainly, evolutionary analysis would hardly be possible without robust classification.

@@

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary deep history: the Palaeolithic. Rooted in French antiquarianism, the development of Palaeolithic systematics, for instance, has been likened to so many 'accidents of history' [@clark2009], but there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units [@reynolds2019]. The use of older typological classifications remains prevalent, despite clear and repeated critiques [e.g. @bisson2000]. More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches. However, theoretical explications of the generative mechanisms, and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell, remain exceptions rather than the rule [@tostevin2013]. Today, a great deal of attention is again being paid to systematics in prehistory, and as Barton and Clark [@barton2021] have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions. Several researchers are tackling classificatory issues with novel and mostly quantitative means [@ivanovait2020; @grove2021; @leplongeon2020]. At the same time, however, the topic remains poorly heeded in archaeological pedagogy [@lyman2021]. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching [@Riede_Walsh_Nowell_Langley_Johannsen_2021], we recommend that, after half a century, Dunnell's *Systematics in Prehistory*—and with it rigorous and replicable ways of classifying material culture—are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [-@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioural archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission,' 'variation,' and 'evolutionary.' Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

```{r sip-citations-of-citing-work, fig.cap="Citations to 'Systematics' over time. Inset shows distributions of citations to works citing 'Systematics', i.e. the degree to which works citing 'Systematics' have themselves been cited. Data collected from Google Scholar in July 2021", fig.width = 5, fig.height = 5}
knitr::include_graphics(here::here("analysis/figures/citations-of-citing-works-over-time.png"))
```

```{r sip-keywords-in-titles, fig.cap="Keywords in titles of works citing 'Systematics'. Inset shows languages of works citing 'Systematics'.", fig.width = 5, fig.height = 6}
knitr::include_graphics(here::here("analysis/figures/languages-and-keywords-in-titles-of-citing-works.png"))
```

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [@Cardillo_Charlin_2018], Mercedes Okumura [@Okumura_Araujo_2014] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. While the work by Cardillo and colleagues [-@Cardillo_Charlin_2018; -@cardillo2015evolution] focuses on phylogenetic signals and material culture diversity across time and space, Okumura and colleagues [-@Okumura_Araujo_2014] are more concerned with the persistence of attributes over long stretches of time. What these works have in common is the use of an explicitly cultural evolutionary rationale. In these studies, the use of paradigmatic classifications is widespread and an overt understanding of the difference between classes (ideational/theoretical) and groups (phenomenological/empirical) evident, underlining the influence of Dunnell’s thinking. Worth noting is that these approaches were embraced chiefly in only two South American countries—Argentina and Brazil—and largely independently of each other. Collectively, these works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields—first and foremost palaeontology—dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

```{r sip-topic-model, fig.cap="Left: Top twenty topics in titles of works citing 'Systematics'. The gamma value indicates their overall abundance, and the topics are labelled with the most heavily weighted words in each topic. Right upper: Clusters of citing works according to topic similarity, with clusters labelled by the most prominent topic. Clusters were computed by Principal Components Analysis of the topic proportions in each citing work, then a t-SNE to reduce dimensionality, and density-based clustering and to identify clusters. Each data point is one document. Right lower: Plot of topics showing coherence and exclusivity metrics for each topic. Each data point is one topic.", fig.width = 5, fig.height = 6}
knitr::include_graphics(here::here("analysis/figures/topic-model-on-titles.jpg"))
```

@@

The influence of *Systematics* has extended well beyond Dunnell's own primary study area of the North American Southeast. For example, the Portuguese translation published in 2007 has received 75 citations, and the Spanish translation 30. At face value these numbers seem low, but when we take into consideration that Binford’s Portuguese translation of _In Pursuit of the Past_, published in 1991, has similarly received a mere 31 citations, we can situate the influence of Dunnell’s work in a more robust perspective. At the same time, in our data we see only a single work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that the book is quite difficult to read, as noted by reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [-@spaulding1974 p. 515-516] described his reaction to the book), and acknowledged by Dunnell in his foreword to the 2002 edition. The low readability of the book has likely limited the accessibility of its contents for readers whose first language is not English, and made it inherently less attractive as a pedagogical resource.

A second limitation of *Systematics* is that it failed to motivate the formation of a distinct research area and community of archaeological systematics. While biology has a Society of Systematic Biologists, and attendant journals such as *Systematic Biology*, *Systematics and Evolution*, and *Systematics and Biodiversity*, archaeology currently has no equivalent focal points for discussions of systematics that transcends specific geographical and chronological concerns. *Systematics* was not a work that served as an engine for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead, these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups. The result of this fragmentation is a proliferation of bespoke classifications that lack applicability across cases. It also inhibits reproducibility, synthesis, and the transfer of ideas and innovations across research communities, which may limit the sustainability of archaeology as a discipline.

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by the New Archaeology on the one hand, while also meeting resistance from deeply entrenched traditional approaches to classification. In addition, the volume also appeared just before personal computers were beginning to have a major impact on archaeological research practice [@clarkQuantificationAmericanArchaeology1982; @aldenderfer1987assessing]. The rigor advocated by Dunnell would have lent itself smoothly to the sorts of computational approaches now increasingly adopted in the discipline. Yet, the approaches of the New Archaeologists dominated the literature in the decade after the publication of *Systematics*, thoroughly eclipsing it as a discipline-defining text. Nevertheless, what makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by economies of personal prestige (Gabriel de Mortillet, 1821--1898, was an archaeologist who published the first widely used classification of the Palaeolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets [cf. @Perreault_2019], the message of *Systematics* will only become more relevant.

@@

We thank Ruth Mace for encouraging us to submit this paper to *Evolutionary Human Sciences*. We also thank Mike O’Brien and one anonymous reviewer for the insightful comments on Dunnell and on our paper.

@@

We declare no conflicts of interest.

@@

The data that support the findings of this study and the R code used to analyse and visualize the data are openly available in the Open Science Framework at <http://doi.org/10.17605/OSF.IO/JBPFW>. Readers may freely access these resources to reproduce all of the figures in the paper.
